---
title: "Introduction to presizepo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to presizepo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(presizepo)
```

## Overview

`presizepo` computes precision-based (confidence-interval width) sample
sizes and standard errors for the log odds ratio in a proportional-odds
(ordinal logistic regression) model. Two predictor types are supported:

* a **binary** group predictor (e.g. treatment vs. control), via
  `se_binary_po()` and `n_precision_binary_po()`;
* a **continuous** predictor, via `se_cont_po()` and
  `n_precision_cont_po()`.

For both, two calculation methods are available via `method`:

* `"whitehead"` - Whitehead's (1993) closed-form delta-method variance
  approximation. Assumes proportional odds and can be
  inaccurate for large effects.
* `"ologit"` (default) - fits a weighted proportional-odds model directly
  to the anticipated distribution and reads the standard error off the
  observed information matrix, following White, Marley-Zagar, Morris, Parmar, Royston & Babiker (2023).

The two methods nearly agree for small effects and diverge as the
anticipated effect size grows.

## Binary predictor

Suppose the anticipated distribution of outcome categories in the
control arm is:

```{r}
pc <- c(0.15, 0.10, 0.10, 0.10, 0.10, 0.45)
```

and we anticipate a proportional odds ratio of 1.7 for the experimental
arm relative to control. Instead of `OR`, the experimental-arm
distribution can also be supplied directly via `pe`, or implied by a
common risk ratio `rr` on the cumulative probabilities; exactly one of
`OR`, `pe`, `rr` must be given.

### Standard error at a fixed sample size

`se_binary_po()` returns the standard error of the log OR, and a
confidence interval for the OR itself, at a given total sample size `n`
(split between arms according to the allocation ratio `r`):

```{r}
se_binary_po(pc, OR = 1.7, n = 200, method = "ologit")
se_binary_po(pc, OR = 1.7, n = 200, method = "whitehead")
```

### Sample size for a target precision

More commonly we want the sample size needed to achieve a target
precision, expressed as the ratio of the upper to lower confidence
limit for the OR (`ratio_UL`). `n_precision_binary_po()` computes this
by fitting once at a large reference sample size and rescaling
analytically (the Fisher information scales linearly with `n`):

```{r}
n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3, method = "ologit")
```

So a sample size of `r n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3)$n`
would be needed to constrain the OR's confidence interval to a
3-fold ratio between its limits.

## Continuous predictor

For a continuous predictor `x`, the anticipated outcome-category
distribution is instead specified at `x = mean(x)` (`p0`), along with the
log OR per unit of `x` (`beta`) and the SD of `x` (`sd_x`):

```{r}
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, method = "ologit")
```

Here `method = "ologit"` integrates out the mixing distribution of `x`
(assumed normal) via Gauss-Hermite quadrature (`ngrid` nodes), jointly
estimating the cutpoints and `beta`; `"whitehead"` instead evaluates a
closed-form variance formula at fixed cutpoints derived from `p0`.

An `R2` argument allows for the predictor's effective information to be
reduced when other covariates explain some of its variance (e.g. when
adjusting for baseline covariates).

`n_precision_cont_po()` gives the sample size for a target CI-width
ratio, expressed for the OR per `delta` units of `x`:

```{r}
n_precision_cont_po(p0, beta = log(1.5), sd_x = 0.5, ratio_UL = 3,
                     delta = 1, method = "ologit")
```


## References

Whitehead, J. (1993). Sample size calculations for ordered categorical
data. *Statistics in Medicine*, 12(24), 2257-2271.

White, I. R., Marley-Zagar, E., Morris, T. P., Parmar, M. K., Royston,
P., & Babiker, A. (2023). A simple software-based approach to sample
size calculation for ordinal outcomes. *Statistical Methods in Medical
Research*. <doi:10.1177/1536867X231162228>
