---
title: "Introduction to presizepo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to presizepo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(presizepo)
```

## Overview

`presizepo` computes precision-based (confidence-interval width) sample
sizes and standard errors for the log odds ratio from a proportional-odds
(ordinal logistic regression) model. Two predictor types are supported:

* a **binary** group predictor (e.g. treatment vs. control), via
  `se_binary_po()` and `n_precision_binary_po()`;
* a **continuous** predictor, via `se_cont_po()` and
  `n_precision_cont_po()`.

For both, two calculation methods are available via `method`:

* `"whitehead"` - Whitehead's (1993) closed-form delta-method variance
  approximation. Assumes proportional odds and can be
  inaccurate for large effects.
* `"ologit"` (default) - fits a weighted proportional-odds model directly
  to the anticipated distribution and reads the standard error off the
  observed information matrix, following White, Marley-Zagar, Morris, Parmar, Royston & Babiker (2023).

The two methods nearly agree for small effects and diverge as the
anticipated effect size grows.

## Binary predictor

Suppose the anticipated distribution of outcome categories in the
control arm is:

```{r}
pc <- c(0.15, 0.10, 0.10, 0.10, 0.10, 0.45)
```

and we anticipate a proportional odds ratio of 1.7 for the experimental
arm relative to control. Instead of `OR`, the experimental-arm
distribution can also be supplied directly via `pe`, or implied by a
common risk ratio `rr` on the cumulative probabilities; exactly one of
`OR`, `pe`, `rr` must be given.

### Standard error at a fixed sample size

`se_binary_po()` returns the standard error of the log OR, and a
confidence interval for the OR itself, at a given total sample size `n`
(split between arms according to the allocation ratio `r`):

```{r}
se_binary_po(pc, OR = 1.7, n = 200, method = "ologit")
se_binary_po(pc, OR = 1.7, n = 200, method = "whitehead")
```

### Sample size for a target precision

More commonly we want the sample size needed to achieve a target
precision, expressed as the ratio of the upper to lower confidence
limit for the OR (`ratio_UL`). `n_precision_binary_po()` computes this
by fitting once at a large reference sample size and rescaling
analytically (the Fisher information scales linearly with `n`):

```{r}
n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3, method = "ologit")
```

So a sample size of `r n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3)$n`
would be needed to constrain the OR's confidence interval to a
3-fold ratio between its limits.

## Continuous predictor

For a continuous predictor `x`, the anticipated outcome-category
distribution is instead specified at `x = mean(x)` (`p0`), along with the
log OR per unit of `x` (`beta`) and the SD of `x` (`sd_x`):

```{r}
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, method = "ologit")
```

Here `method = "ologit"` integrates out the mixing distribution of `x`
(assumed normal) via Gauss-Hermite quadrature (`ngrid` nodes), jointly
estimating the cutpoints and `beta`; `"whitehead"` instead evaluates a
closed-form variance formula at fixed cutpoints derived from `p0`.

An `R2` argument allows for the predictor's effective information to be
reduced when other covariates explain some of its variance (e.g. when
adjusting for baseline covariates).

### Effect size parameterization: `beta` vs `beta_sd`

The effect size can be specified in two ways:

* `beta`: log OR per **raw unit** of `x`
* `beta_sd`: log OR per **SD** of `x`

These are equivalent ways of specifying the same effect. For example, with
`sd_x = 0.5`, a log OR per raw unit of `log(1.5)` is equivalent to a log OR
per SD of `log(1.5) * 0.5`:

```{r}
# beta: log OR per raw unit
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, method = "ologit")

# beta_sd: log OR per SD (equivalent when sd_x = 0.5)
se_cont_po(p0, beta_sd = log(1.5) * 0.5, sd_x = 0.5, n = 200, method = "ologit")
```

### Change parameterization: `delta` vs `delta_sd`

Similarly, the reported change in `x` can be expressed in two ways:

* `delta`: change in **raw units** of `x`
* `delta_sd`: change in **SDs** of `x`

These control only what change in `x` the *output* is expressed over; they 
don't change the underlying effect size. For example, with `sd_x = 0.5`:

```{r}
# delta: OR for a 1-unit change in x
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, delta = 1, method = "ologit")

# delta_sd: OR for a 1-SD change in x (= 0.5 raw units when sd_x = 0.5)
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, delta_sd = 1, method = "ologit")
```

Note that the reported log ORs differ: with `delta = 0.5` we get the same
result as `delta_sd = 1`:

```{r}
# delta = 0.5 raw units vs delta_sd = 1 SD (equivalent when sd_x = 0.5)
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, delta = 0.5, method = "ologit")
se_cont_po(p0, beta = log(1.5), sd_x = 0.5, n = 200, delta_sd = 1, method = "ologit")
```

### Sample size for a target precision

`n_precision_cont_po()` gives the sample size for a target CI-width
ratio, expressed for the OR per `delta` units of `x`. It supports the same
`beta`/`beta_sd` and `delta`/`delta_sd` options:

```{r}
# Using beta and delta
n_precision_cont_po(p0, beta = log(1.5), sd_x = 0.5, ratio_UL = 3,
                     delta = 1, method = "ologit")

# Using beta_sd and delta_sd (specifying effect and change per SD)
n_precision_cont_po(p0, beta_sd = log(1.5) * 0.5, sd_x = 0.5, ratio_UL = 3,
                     delta_sd = 1, method = "ologit")
```

## Simulation validation

To validate that the theoretical standard errors match empirical results, we can
simulate data from a proportional-odds model and compare the empirical standard
error with the one predicted by `se_cont_po()`.

```{r}
# Set up simulation parameters
set.seed(1234)
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
beta <- log(1.5)
sd_x <- 0.5
n_sim <- 1000  # number of simulations

# Predicted SE from presizepo
pred_se <- se_cont_po(p0, beta = beta, sd_x = sd_x, n = 200, method = "ologit")
cat("Predicted SE (from presizepo):", pred_se[["SE"]], "\n")

# Simulation
set.seed(1234)
beta_hats <- numeric(n_sim)

for (i in seq_len(n_sim)) {
  # Generate predictor x from N(0, sd_x^2)
  x <- rnorm(1000, mean = 0, sd = sd_x)
  
  # Generate ordinal outcome from proportional-odds model
  eta <- beta * x
  p_cumsum <- stats::plogis(stats::qlogis(cumsum(p0)[-length(p0)]) - eta)
  p_mat <- cbind(p_cumsum, 1)
  p_mat <- cbind(0, p_mat)
  p_probs <- t(apply(p_mat, 1, diff))
  
  y <- apply(p_probs, 1, function(p) sample(seq_along(p0), size = 1, prob = p))
  
  # Fit proportional-odds model
  fit <- suppressWarnings(
    MASS::polr(factor(y) ~ x, method = "logistic", Hess = TRUE)
  )
  
  # Extract beta estimate (slope coefficient)
  beta_hats[i] <- unname(coef(fit))
}

# Empirical SE
emp_se <- sd(beta_hats)
cat("Empirical SE (from simulation):", emp_se, "\n")
cat("Ratio (empirical / predicted):", emp_se / pred_se[["SE"]], "\n")
```

The empirical SE from the simulation should closely match the predicted SE,
validating the theoretical calculations. The ratio should be close to 1.

We can also visualize the distribution of beta estimates:

```{r}
hist(beta_hats, breaks = 30, main = "Distribution of beta estimates from simulation",
     xlab = "Estimated beta", freq = FALSE)
abline(v = beta, col = "red", lwd = 2, label = "True beta")
abline(v = mean(beta_hats), col = "blue", lwd = 2, label = "Mean of estimates")
legend("topright", c("True beta", "Mean of estimates"), col = c("red", "blue"), lwd = 2)
```

## References

Whitehead, J. (1993). Sample size calculations for ordered categorical
data. *Statistics in Medicine*, 12(24), 2257-2271.

White, I. R., Marley-Zagar, E., Morris, T. P., Parmar, M. K., Royston,
P., & Babiker, A. (2023). A simple software-based approach to sample
size calculation for ordinal outcomes. *Statistical Methods in Medical
Research*. <doi:10.1177/1536867X231162228>
