---
title: "Introduction to presizepo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to presizepo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(presizepo)
```

## Overview

`presizepo` is precision-based (confidence-interval width) sample
size calculation approach for the log common odds ratio (cOR) from a proportional-odds 
model. Two predictor types are supported:

* a **binary** predictor (e.g. treatment vs control), via
  `se_binary_po()` and `n_precision_binary_po()`;
* a **continuous** predictor, via `se_cont_po()` and
  `n_precision_cont_po()`.

For both, two calculation methods are available via `method`:

* `"whitehead"` - Whitehead's (1993) formula.
* `"ologit"` (default) - fits a weighted proportional-odds model directly
  to the anticipated distribution and reads the standard error off the
  observed information matrix, following White, Marley-Zagar, Morris, Parmar, Royston & Babiker (2023).

The two methods are similar for small effects and diverge as the
anticipated effect size grows.

`presizepo` offers two approaches for calculating the standard error of the log odds ratio.

### Whitehead's (1993) Formula

Whitehead (1993) originally derived his formula for **hypothesis testing** sample size calculations with specified probability of Type I error ($\alpha$) and power ($1 - \beta$). In this setting the sample size can be generally formulated as (see Julious (2004))

$$n = \frac{V(z_{\alpha/2} + z_\beta)^2}{d^2},$$
where $d$ is the anticipated effect size and $V$ is the variance of the test statistic. Whitehead (1993) showed that for a binary predictor and allocation ratio $r$ (treatment:control),

$$n = \frac{3(r+1)^2(z_{\alpha/2} + z_\beta)^2}{rd^2(1 - \sum_{k} \bar{p}_k^3)},$$
where $d$ is the anticipated log OR and $\bar{p}_k$ is the average probability for outcome category $k$ across treatment arms, weighted by allocation $$\bar{p}_k = \frac{p_{c,k} + r \cdot p_{t,k}}{1 + r}.$$ Thus, the standard error for the anticipated log OR $d$ is

$$\text{SE}(d) = \sqrt{\frac{3(r+1)^2}{rn(1 - \sum_{k} \bar{p}_k^3)}}.$$
From this relationship, one can solve for the sample size needed to achieve a target confidence interval width.

For a continuous predictor $x \sim N(0, \sigma_x^2)$ one can adapt Whitehead's formula to:

$$\text{SE}(d) = \sqrt{\frac{3(r+1)^2}{rn\sigma_x^2 (1 - \sum_{k} \bar{p}_k^3) (1 - R^2)}},$$

where $(1 - R^2)$ adjusts for information loss when the predictor is correlated with other covariates in the model.

**Implementation details**: For a continuous predictor, the average probabilities $\bar{p}_k$ are computed by integrating the outcome probabilities across the distribution of $x$ using a fixed quantile grid approach with 2000 points from $x \sim N(0, \sigma_x^2)$.

### White et al. (2023) Method: "ologit"

White et al. (2023) propose a more direct approach that avoids delta-method approximation by fitting a weighted proportional-odds model to the anticipated data distribution and extracting the standard error from the observed information matrix.

For binary predictors the algorithm is:

1. **Construct anticipated data**: Create a weighted frequency table from the anticipated outcome category probabilities at each treatment level (control and treatment arms).
2. **Fit weighted proportional-odds model**: Fit a proportional odds model to this table, with weights proportional to arm size and allocation ratio.
3. **Extract SE**: Read the standard error of the treatment coefficient ($\beta$) directly from the inverse Hessian matrix (observed information matrix).

For continuous $x \sim N(0, \sigma_x^2)$ one adapts to:

1. **Integrate over the predictor distribution**: Use Gauss-Hermite quadrature with $k$ nodes (default $k=15$) to approximate the mixing distribution of $x$.
2. **Construct weighted pseudo-data**: At each quadrature node $x_j$ with weight $w_j$:
   - Compute outcome category probabilities: $p_{k}(x_j) = \text{Pr}(Y = k | X = x_j)$
   - Weight by: $w_{k,j} = p_k(x_j) \cdot w_j \cdot n \cdot (1 - R^2)$
3. **Fit weighted model**: Fit a weighted proportional-odds model to the pooled pseudo-observations across all nodes and both treatment arms (if applicable).
4. **Extract SE**: Read the standard error of the $\beta$ coefficient from the inverse Hessian.

The Gauss-Hermite nodes and weights are scaled to match $x \sim N(0, \sigma_x^2)$ via the transformation: $x_j = \sqrt{2} \sigma_x \cdot n_j$, where $n_j$ are the standard Hermite nodes.

## Examples

### Binary predictor

Suppose the anticipated distribution of outcome categories in the
control arm is:

```{r}
pc <- c(0.15, 0.10, 0.10, 0.10, 0.10, 0.45)
```

and we anticipate a cOR of 1.7 for the experimental
arm relative to control.

#### Standard error at a fixed sample size

`se_binary_po()` returns the standard error of the log cOR, and a
confidence interval for the cOR itself, at a given total sample size `n`:

```{r}
se_binary_po(pc, OR = 1.7, n = 200, method = "ologit")
se_binary_po(pc, OR = 1.7, n = 200, method = "whitehead")
```

#### Sample size for a target precision

More commonly we want the sample size needed to achieve a target
precision, expressed as the ratio of the upper to lower confidence
limit for the OR (`ratio_UL`). `n_precision_binary_po()` computes this, here for
`ratio_UL=3`.

```{r}
n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3, method = "ologit")
```

So a sample size of `r n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3)$n`
would be needed to constrain the cOR's confidence interval to a
3-fold ratio between its limits.

### Continuous predictor

For a continuous predictor `x`, the anticipated outcome-category
distribution is instead specified at `x = 0`, along with the cOR per unit of `x` (`OR`) and the SD of `x` (`sd_x`):

```{r}
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, method = "ologit")
```

An `R2` argument allows for the predictor's effective information to be
reduced when other covariates explain some of its variance (e.g. when
adjusting for baseline covariates). For example, if other covariates explain
30% of the variance of `x`, set `R2 = 0.3`:

```{r}
# Without adjustment for other covariates (R2 = 0)
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, R2 = 0, method = "ologit")

# With other covariates explaining 30% of variance of x (R2 = 0.3)
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, R2 = 0.3, method = "ologit")
```

Notice that the SE increases when `R2 = 0.3` (and correspondingly the confidence 
interval widens), reflecting the loss of information due to confounding or 
correlation with other covariates.

#### Effect size parameterization: `OR` vs `OR_sd`

The effect size can be specified in two ways:

* `OR`: cOR per **raw unit** of `x`
* `OR_sd`: cOR per **SD** of `x`

These are equivalent ways of specifying the same effect. For example, with
`sd_x = 0.5`, a cOR per raw unit of `log(1.5)` is equivalent to a cOR
per SD of `log(sqrt(1.5)`:

```{r}
# OR: log OR per raw unit
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, method = "ologit")

# OR_sd: log OR per SD (equivalent when sd_x = 0.5)
se_cont_po(p0, OR_sd = sqrt(1.5), sd_x = 0.5, n = 200, method = "ologit")
```

#### Change parameterization: `delta` vs `delta_sd`

Similarly, the reported change in `x` can be expressed in two ways:

* `delta`: change in **raw units** of `x`
* `delta_sd`: change in **SDs** of `x`

For example, with `sd_x = 0.5`:

```{r}
# delta = 0.5 raw units vs delta_sd = 1 SD (equivalent when sd_x = 0.5)
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, delta = 0.5, method = "ologit")
se_cont_po(p0, OR = 1.5, sd_x = 0.5, n = 200, delta_sd = 1, method = "ologit")
```

In the first example, we express the change in `x` in a raw unit increase of $0.5$. This is the same, as when we express it by a 'one SD' increase `delta_sd=1`.

#### Sample size for a target precision

`n_precision_cont_po()` gives the sample size for a target CI-width
ratio, expressed for the OR per `delta` units of `x`. It supports the same
`OR`/`OR_sd` and `delta`/`delta_sd` options:

```{r}
# Using OR and delta
n_precision_cont_po(p0, OR = 1.5, sd_x = 0.5, ratio_UL = 3,
                     delta = 1, method = "ologit")

# Using OR_sd and delta_sd (specifying effect and change per SD)
n_precision_cont_po(p0, OR_sd = sqrt(1.5), sd_x = 0.5, ratio_UL = 3,
                     delta_sd = 1, method = "ologit")
```

## Simulation validation

To validate that the theoretical standard errors match empirical results, we can
simulate data from a proportional-odds model and compare the empirical standard
error with the one predicted by `se_cont_po()`.

```{r, message=FALSE}
library(MASS)

# Set up simulation parameters
set.seed(1234)
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
OR <- 1.5  # odds ratio per unit of x
beta <- log(OR)  # log OR for data generation
sd_x <- 0.5
n_sim <- 500  # number of simulations
n_obs <- 200  # observations per simulation

# Predicted SE from presizepo
pred_se <- se_cont_po(p0, OR = OR, sd_x = sd_x, n = n_obs, method = "ologit")
cat("Predicted SE (from presizepo):", pred_se[["se_log_OR"]], "\n")

# Simulation: fit the model many times and collect beta estimates
set.seed(1234)
beta_hats <- numeric(n_sim)

for (i in seq_len(n_sim)) {
  # Generate predictor x from N(0, sd_x^2)
  x <- rnorm(n_obs, mean = 0, sd = sd_x)
  
  # Generate ordinal outcome from proportional-odds model
  # Compute cumulative probabilities: P(Y <= k)
  cutpoints <- qlogis(cumsum(p0)[-length(p0)])
  eta <- beta * x
  p_cumulative <- plogis(outer(cutpoints, eta, "-"))
  
  # Convert to category probabilities: P(Y = k)
  p_cat <- rbind(p_cumulative[1, ], 
                 diff(p_cumulative, margin = 1),
                 1 - p_cumulative[nrow(p_cumulative), ])
  
  # Generate outcomes: sample category for each observation
  y <- apply(p_cat, 2, function(p) sample(seq_along(p0), size = 1, prob = p))
  
  # Fit proportional-odds model
  fit <- suppressWarnings(
    polr(factor(y) ~ x, method = "logistic", Hess = TRUE)
  )
  
  # Extract coeff
  beta_hats[i] <- unname(coef(fit))
}

# Empirical SE
emp_se <- sd(beta_hats)
cat("Empirical SE (from simulation):", emp_se, "\n")
cat("Ratio (empirical / presizepo):", emp_se / pred_se[["se_log_OR"]], "\n")
```

The empirical SE from the simulation should closely match the predicted SE,
validating the theoretical calculations. The ratio should be close to 1.

Distribution of beta estimates:

```{r, echo=FALSE}
hist(beta_hats, breaks = 30, main = "Distribution of coefficients",
     xlab = "Estimated coefficients", freq = FALSE)
abline(v = beta, col = "red", lwd = 2)
abline(v = mean(beta_hats), col = "blue", lwd = 2)
legend("topright", c("presizepo", "Mean of estimates"), col = c("red", "blue"), lwd = 2)
```

## References

Julious, S.A. (2004) Tutorial in biostatistics - Sample sizes for clinical trials with Normal
data. *Statistics in Medicine*, 23 (12). pp. 1921-1986.

Whitehead, J. (1993). Sample size calculations for ordered categorical
data. *Statistics in Medicine*, 12(24), 2257-2271.

White, I. R., Marley-Zagar, E., Morris, T. P., Parmar, M. K., Royston,
P., & Babiker, A. (2023). A simple software-based approach to sample
size calculation for ordinal outcomes. *Statistical Methods in Medical
Research*. <doi:10.1177/1536867X231162228>
