---
title: "Introduction to presizepo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to presizepo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(presizepo)
```

## Overview

`presizepo` is precision-based (confidence-interval width) sample
size calculation approach for the log common odds ratio (cOR) from a proportional-odds 
model. Two predictor types are supported:

* a **binary** predictor (e.g. treatment vs control), via
  `se_binary_po()` and `n_precision_binary_po()`;
* a **continuous** predictor, via `se_cont_po()` and
  `n_precision_cont_po()`.

For both, two calculation methods are available via `method`:

* `"whitehead"` - Whitehead's (1993) formula.
* `"ologit"` (default) - fits a weighted proportional-odds model directly
  to the anticipated distribution and reads the standard error off the
  observed information matrix, following White, Marley-Zagar, Morris, Parmar, Royston & Babiker (2023).

The two methods are similar for small effects and diverge as the
anticipated effect size grows.

## Binary predictor

Suppose the anticipated distribution of outcome categories in the
control arm is:

```{r}
pc <- c(0.15, 0.10, 0.10, 0.10, 0.10, 0.45)
```

and we anticipate a cOR of 1.7 for the experimental
arm relative to control.

### Standard error at a fixed sample size

`se_binary_po()` returns the standard error of the log cOR, and a
confidence interval for the cOR itself, at a given total sample size `n`:

```{r}
se_binary_po(pc, OR = 1.7, n = 200, method = "ologit")
se_binary_po(pc, OR = 1.7, n = 200, method = "whitehead")
```

### Sample size for a target precision

More commonly we want the sample size needed to achieve a target
precision, expressed as the ratio of the upper to lower confidence
limit for the OR (`ratio_UL`). `n_precision_binary_po()` computes this.

```{r}
n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3, method = "ologit")
```

So a sample size of `r n_precision_binary_po(pc, OR = 1.7, ratio_UL = 3)$n`
would be needed to constrain the cOR's confidence interval to a
3-fold ratio between its limits.

## Continuous predictor

For a continuous predictor `x`, the anticipated outcome-category
distribution is instead specified at `x = mean(x)` (`p0`), along with the
log cOR per unit of `x` (`beta`) and the SD of `x` (`sd_x`):

```{r}
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, method = "ologit")
```

An `R2` argument allows for the predictor's effective information to be
reduced when other covariates explain some of its variance (e.g. when
adjusting for baseline covariates). For example, if other covariates explain
30% of the variance of `x`, set `R2 = 0.3`:

```{r}
# Without adjustment for other covariates (R2 = 0)
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, R2 = 0, method = "ologit")

# With other covariates explaining 30% of variance of x (R2 = 0.3)
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, R2 = 0.3, method = "ologit")
```

Notice that the SE increases when `R2 = 0.3` (and correspondingly the confidence 
interval widens), reflecting the loss of information due to confounding or 
correlation with other covariates.

### Effect size parameterization: `OR` vs `OR_sd`

The effect size can be specified in two ways (note: these are log OR values):

* `OR`: log OR per **raw unit** of `x`
* `OR_sd`: log OR per **SD** of `x`

These are equivalent ways of specifying the same effect. For example, with
`sd_x = 0.5`, a log OR per raw unit of `log(1.5)` is equivalent to a log OR
per SD of `log(1.5) * 0.5`:

```{r}
# OR: log OR per raw unit
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, method = "ologit")

# OR_sd: log OR per SD (equivalent when sd_x = 0.5)
se_cont_po(p0, OR_sd = log(1.5) * 0.5, sd_x = 0.5, n = 200, method = "ologit")
```

### Change parameterization: `delta` vs `delta_sd`

Similarly, the reported change in `x` can be expressed in two ways:

* `delta`: change in **raw units** of `x`
* `delta_sd`: change in **SDs** of `x`

These control only what change in `x` the *output* is expressed over; they 
don't change the underlying effect size. For example, with `sd_x = 0.5`:

```{r}
# delta: OR for a 1-unit change in x
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, delta = 1, method = "ologit")

# delta_sd: OR for a 1-SD change in x (= 0.5 raw units when sd_x = 0.5)
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, delta_sd = 1, method = "ologit")
```

Note that the reported log ORs differ: with `delta = 0.5` we get the same
result as `delta_sd = 1`:

```{r}
# delta = 0.5 raw units vs delta_sd = 1 SD (equivalent when sd_x = 0.5)
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, delta = 0.5, method = "ologit")
se_cont_po(p0, OR = log(1.5), sd_x = 0.5, n = 200, delta_sd = 1, method = "ologit")
```

### Sample size for a target precision

`n_precision_cont_po()` gives the sample size for a target CI-width
ratio, expressed for the OR per `delta` units of `x`. It supports the same
`OR`/`OR_sd` and `delta`/`delta_sd` options:

```{r}
# Using OR and delta
n_precision_cont_po(p0, OR = log(1.5), sd_x = 0.5, ratio_UL = 3,
                     delta = 1, method = "ologit")

# Using OR_sd and delta_sd (specifying effect and change per SD)
n_precision_cont_po(p0, OR_sd = log(1.5) * 0.5, sd_x = 0.5, ratio_UL = 3,
                     delta_sd = 1, method = "ologit")
```

## Simulation validation

To validate that the theoretical standard errors match empirical results, we can
simulate data from a proportional-odds model and compare the empirical standard
error with the one predicted by `se_cont_po()`.

```{r, message=FALSE}
library(MASS)

# Set up simulation parameters
set.seed(1234)
p0 <- c(0.10, 0.15, 0.15, 0.20, 0.20, 0.10, 0.10)
beta <- log(1.5)
sd_x <- 0.5
n_sim <- 500  # number of simulations
n_obs <- 200  # observations per simulation

# Predicted SE from presizepo
pred_se <- se_cont_po(p0, OR = beta, sd_x = sd_x, n = n_obs, method = "ologit")
cat("Predicted SE (from presizepo):", pred_se[["se_log_OR"]], "\n")

# Simulation: fit the model many times and collect beta estimates
set.seed(1234)
beta_hats <- numeric(n_sim)

for (i in seq_len(n_sim)) {
  # Generate predictor x from N(0, sd_x^2)
  x <- rnorm(n_obs, mean = 0, sd = sd_x)
  
  # Generate ordinal outcome from proportional-odds model
  # Compute cumulative probabilities: P(Y <= k)
  cutpoints <- qlogis(cumsum(p0)[-length(p0)])
  eta <- beta * x
  p_cumulative <- plogis(outer(cutpoints, eta, "-"))
  
  # Convert to category probabilities: P(Y = k)
  p_cat <- rbind(p_cumulative[1, ], 
                 diff(p_cumulative, margin = 1),
                 1 - p_cumulative[nrow(p_cumulative), ])
  
  # Generate outcomes: sample category for each observation
  y <- apply(p_cat, 2, function(p) sample(seq_along(p0), size = 1, prob = p))
  
  # Fit proportional-odds model
  fit <- suppressWarnings(
    polr(factor(y) ~ x, method = "logistic", Hess = TRUE)
  )
  
  # Extract beta estimate (slope coefficient)
  beta_hats[i] <- unname(coef(fit))
}

# Empirical SE
emp_se <- sd(beta_hats)
cat("Empirical SE (from simulation):", emp_se, "\n")
cat("Ratio (empirical / predicted):", emp_se / pred_se[["se_log_OR"]], "\n")
```

The empirical SE from the simulation should closely match the predicted SE,
validating the theoretical calculations. The ratio should be close to 1.

We can also visualize the distribution of beta estimates:

```{r}
hist(beta_hats, breaks = 30, main = "Distribution of beta estimates from simulation",
     xlab = "Estimated beta", freq = FALSE)
abline(v = beta, col = "red", lwd = 2)
abline(v = mean(beta_hats), col = "blue", lwd = 2)
legend("topright", c("True beta", "Mean of estimates"), col = c("red", "blue"), lwd = 2)
```

## References

Whitehead, J. (1993). Sample size calculations for ordered categorical
data. *Statistics in Medicine*, 12(24), 2257-2271.

White, I. R., Marley-Zagar, E., Morris, T. P., Parmar, M. K., Royston,
P., & Babiker, A. (2023). A simple software-based approach to sample
size calculation for ordinal outcomes. *Statistical Methods in Medical
Research*. <doi:10.1177/1536867X231162228>
